Hire Ed: Job Market Dynamics for Tenure-Track Faculty Positions in Archaeology

### 2024-11-05 analysis/paper/article.qmd
discussion and conclusion

In addition to a dip in job ads, our results show that 2019-2021 was a major inflection point in the popularity of topics and regions in job ads. Indigenous and historical archaeology and the Americas strongly increase in popularity at this time, while archaeological science, complex societies, the Mediterranean and Near East and show declines. Similarly, the number of job ads with a geographic focus on the Americas and Africa peaks during this time. These shifts in the topics and geographic foci of job ads were likely been influenced by broader cultural movements such as Black Lives Matter, protests about racial injustice, and efforts to amplify Indigenous voices. COVID-19 negatively impacted Black, American Indian and Hispanic communities with significantly higher infection and morbidity rates, drawing attention to racial and socio-economic inequality in the US. The Black Lives Matter movement, dating back to 2013, intersected profoundly with the pandemic with the murder of George Floyd by a police officer in May 2020, three months after the World Health Organization declared COVID-19 to be a pandemic. Mass protests objecting to Floyd's murder generated widespread concern about racial inequities and stimulated a broad interest in addressing systemic racial injustice. Our data suggest that US universities participated in this movement by adjusting their hiring plans to prioritize recruiting archaeologists working on topics relevant to Black and Indigenous communities. Many universities may have hoped to hire Black and Indigenous archaeologists as part of their effort to tackle systemic racism. However as it is illegal in the US for universities use race or ethnicity as a primary factor in hiring (due to the 1964 Civil Rights Act), universities appear to have tailored the content of the job ads to focus on topics where they expect Black and Indigenous researchers to be most numerous, ensuring that these researchers would be well-represented in the pool of applicants.

This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. We found a long-term decline in topics such as the archaeology of complex societies and archaeological science, Mesoamerica and South America as a geographic focus, and methods relating to landscape archaeology. These trends are harder to explain as we cannot link their start to a discrete moment like COVID-19. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, relational perspectives in combination with human subjectivity and cultural framing, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism. This might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical methods for compositional analysis. Other relevant factors may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation, and an increasing emphasis on more job-oriented degrees that prepare students for working in cultural heritage management rather than advanced research. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past, increased interest in climate change, environmental sustainability, and resilience which may have shifted attention away from the study of monumental architecture, elite societies, political hierarchies and state systems.

Our data on the requirements for applicants shows that application complexity has gradually increased over time. This trend is especially pronounced for Assistant Professor positions, which make more demands on applicants than Associate and Full Professor positions. We saw a growing demand for research and teaching statements, consistent with previous work. Demand for diversity statements shows a unique trajectory, peaking in 2020-2021 and declining into the present. This may relate to the intersecting concerns emerging during the COVID-19 pandemic, which may have declined in job ads as the pandemic moves into the past, in addition to the debates surrounding the experiment with diversity statements in hiring at some UC campuses. The decline in recent years for names of recommenders in the initial application may be a response to criticisms of the burden on the applicant of preparing numerous complex job applications [e.g. @dennisWorstAnthroJob2022]. In recognition of this burden, not only on applicants but also on colleagues writing letters of recommendation over and over to support applicants, many hiring committees now follow the recommendations of @dennisWorstAnthroJob2022 in only requesting names and letters of recommendation at later stages of the hiring process, if at all.

>>>>

In addition to a dip in job ads, our results show that 2019-2021 was a major inflection point in the popularity of topics and regions in job ads. Indigenous and historical archaeology and the Americas strongly increase in popularity at this time, while archaeological science, complex societies, the Mediterranean and Near East and show declines. Similarly, the number of job ads with a geographic focus on the Americas and Africa peaks during this time. These shifts in the topics and geographic foci of job ads were likely influenced by broader cultural movements such as Black Lives Matter, protests about racial injustice, and efforts to amplify Indigenous voices. COVID-19 negatively impacted Black, American Indian and Hispanic communities with significantly higher infection and morbidity rates, drawing attention to racial and socio-economic inequality in the US. The Black Lives Matter movement, dating back to 2013, intersected profoundly with the pandemic with the murder of George Floyd in Minneapolis by a police officer in May 2020, three months after the World Health Organization declared COVID-19 to be a pandemic. Mass protests objecting to Floyd's murder generated widespread concern about racial inequities and stimulated a broad interest in addressing systemic racial injustice. Our data suggest that US universities participated in this movement by adjusting their hiring plans to prioritize recruiting archaeologists working on topics relevant to Black and Indigenous communities. Many universities may have hoped to hire Black and Indigenous archaeologists as part of their effort to tackle systemic racism. However as it is illegal in the US for universities use race or ethnicity as a primary factor in hiring (due to the 1964 Civil Rights Act which prohibits hiring based solely on race or ethnicity), universities appear to have tailored the content of the job ads to focus on topics where they expect Black and Indigenous researchers to be most numerous, in an effort to ensure that these researchers would be well-represented in the pool of applicants.

This striking change in topics and geographic foci of 2019-2021 occurred against a backdrop of several longer-term trends. We found a long-term decline in topics such as the archaeology of complex societies and archaeological science, Mesoamerica and South America as a geographic focus, and methods relating to landscape archaeology. These trends are harder to explain as we cannot link their start to a moment like COVID-19. We might speculate that a growing preference for archaeological approaches that privilege agency-driven, relational perspectives, in combination with human subjectivity and cultural framing, is one enduring legacy of debates in the 1980s and 90s about processualism versus post-processualism. This might explain why archaeological science is showing a decline, as demand for methods for analyzing artefact materiality, ontology, and power displace physical laboratory methods for compositional analysis. Other relevant factors may include increasing difficulty of obtaining research funding to support archaeological science research, such as laboratory facilities and instrumentation, and an increasing emphasis on more job-oriented degrees that prepare students for working in cultural heritage management rather than advanced research. A decline in interest in the archaeology of complex societies may reflect several themes that intersect with broader social changes, such as growing interest in Indigenous and non-state actors in the past, increased interest in climate change, environmental sustainability, and resilience, which may have shifted attention away from the study of monumental architecture, elite societies, political hierarchies and state systems.

Our data on the requirements for applicants support prior findings that application complexity has gradually increased over time. This trend is especially pronounced for Assistant Professor positions, which make more demands on applicants than Associate and Full Professor positions. We saw a growing demand for research and teaching statements, consistent with previous work. Demand for diversity statements shows a unique trajectory, peaking in 2020-2021 and declining into the present. This may relate to the intersecting concerns emerging during the COVID-19 pandemic, which may have declined in job ads as the most urgent time of the pandemic moves into the past. Another factor here may be the debates surrounding the experiment with diversity statements in hiring at some UC campuses during 2016-2022. Universities that previously required a diversity statement dropped that requirement after 2022 [@guidenAreDiversityStatements]. In our data we observed one univeristy . The decline in recent years for names of recommenders in the initial application may be a response to criticisms of the burden on the applicant of preparing numerous complex job applications [e.g. @dennisWorstAnthroJob2022]. In recognition of this burden, not only on applicants but also on colleagues writing letters of recommendation over and over to support applicants, many hiring committees now follow the recommendations of @dennisWorstAnthroJob2022 in only requesting names and letters of recommendation at later stages of the hiring process, if at all.
